---
title: "Analysing simulations"
author: "Janetta E. Skarp"
date: "28 June 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparing the full dataset for analysis

In my forecasting folder, I have a folder called simulations, in which I hold all of my simulation data. Each simulation has its own folder, and from these folders I want a .csv file called "full_proj_metrics.csv". I want to put all of these csvs together to make a massive table.

```{r collate_csvs}
library(data.table)

setwd("/home/evelina/Development/forecasting/simulations/ebola_test_3/")

# list all files in directory named full_proj_metrics.csv
output_files <- list.files("/home/evelina/Development/forecasting/simulations/ebola_test_3/", pattern = "full_proj_metrics.csv", 
                           full.names = TRUE, recursive = TRUE)

# read and row bind all data sets
combo_table <- rbindlist(lapply(output_files, fread))
```

## Visualisation of metrics

To see how the collated simulations' metrics are distributed, I will plot them.

### Reliability

The value that I really care about here is the value for uniformity for each projection window (calculated by the chi-squared test) but currently I have daily CDFs. I want to collate the data so that I have the data for each projection projection day for each projection window. Therefore I care about days_since_data, proj_window_number, and cali_window_size.

```{r reliability}
library(dplyr)
library(ggplot2)

# subset the big combo table to calculate daily reliability and plot
cali_window <- unique(combo_table$cali_window_size)

for (i in c(cali_window)) {
  cali_table <- dplyr::filter(combo_table, cali_window_size == i)
  proj_window <- unique(cali_table$proj_window_no)
  cali_table$proj_window_no <- as.factor(cali_table$proj_window_no)
  proj_plot <- ggplot(cali_table, aes(x = proj_window_no, y = reliability)) + 
                 geom_violin(trim = TRUE, fill = "lightskyblue") +
                 coord_cartesian(ylim = c(0.75, 1)) +
                 geom_boxplot(width = 0.05) +
                 facet_wrap(~ pred_type) +
                 labs(title = paste("Reliability for calibration window of size", i, sep = " "))
  print(proj_plot)
}
```

### Sharpness

Here I care about the daily value of sharpness - ie. how spread out the predictions are. 

Sharpness lowers with increasing days, but if the outbreak dies out early, the projections will be at 0 at which point sharpness is 1 because there are no cases. Note that because sharpness looks at medians, it does not do a good job at taking outliers into account, that's what RMSE is for.

```{r sharpness}
library(dplyr)
library(ggplot2)

# subset the big combo table to calculate daily reliability and plot
cali_window <- unique(combo_table$cali_window_size)

for (i in c(cali_window)) {
  cali_table <- dplyr::filter(combo_table, cali_window_size == i)
  proj_window <- unique(cali_table$proj_window_no)
  cali_table$proj_window_no <- as.factor(cali_table$proj_window_no)
  proj_plot <- ggplot(cali_table, aes(x = proj_window_no, y = sharpness)) + 
                 geom_violin(trim = TRUE, fill = "lightskyblue") +
                 coord_cartesian(ylim = c(0, 1)) +
                 geom_boxplot(width = 0.05) +
                 facet_wrap(~ pred_type) +
                 labs(title = paste("Sharpness for calibration window of size", i, sep = " "))
  print(proj_plot)
}
# Seeing why there are no violin plots for this combination - there is data but I guess it's just not being plotted..?
# combo_table_50 <- subset(combo_table, cali_window_size == 50)
# 
# combo_table_20_3 <- subset(combo_table, cali_window_size == 20 &
#                         pred_type == 3)
```

### Bias

This says whether your projections are mostly above or below the true value. The bias score can also be used to compare different projection windows - it might need to be done for windows of 10 though.

```{r bias}
library(dplyr)
library(ggplot2)

# subset the big combo table to calculate daily reliability and plot
cali_window <- unique(combo_table$cali_window_size)

for (i in c(cali_window)) {
  cali_table <- dplyr::filter(combo_table, cali_window_size == i)
  proj_window <- unique(cali_table$proj_window_no)
  cali_table$proj_window_no <- as.factor(cali_table$proj_window_no)
  proj_plot <- ggplot(cali_table, aes(x = proj_window_no, y = bias)) + 
                 geom_violin(trim = TRUE, fill = "lightskyblue") +
                 coord_cartesian(ylim = c(-1, 1)) +
                 geom_boxplot(width = 0.05) +
                 facet_wrap(~ pred_type) +
                 labs(title = paste("Bias for calibration window of size", i, sep = " "))
  print(proj_plot)
}
```

### RMSE

```{r rmse}
library(dplyr)
library(ggplot2)

# subset the big combo table to calculate daily reliability and plot
cali_window <- unique(combo_table$cali_window_size)

for (i in c(cali_window)) {
  cali_table <- dplyr::filter(combo_table, cali_window_size == i)
  proj_window <- unique(cali_table$proj_window_no)
  cali_table$proj_window_no <- as.factor(cali_table$proj_window_no)
  proj_plot <- ggplot(cali_table, aes(x = proj_window_no, y = rmse)) + 
                 geom_violin(trim = TRUE, fill = "lightskyblue") +
                 geom_boxplot(width = 0.05) +
                 facet_wrap(~ pred_type) +
                 # coord_cartesian(ylim = c(0, 300)) +
                 labs(title = paste("RMSE for calibration window of size", i, sep = " "))
  print(proj_plot)
}
```